---
title: "Data structures in R"
author: "Eralda Gjika"
date: "February 2022"
output:
  word_document: default
  html_notebook: default
subtitle: GGPLOT and Esquisse library
---


### Get started with R Markdown 

Reference:
https://rmarkdown.rstudio.com/articles_intro.html 
https://rmarkdown.rstudio.com/lesson-1.html

## Data sctructures in R
Most of the data in R are saved in a dataframe format or tibble format . This will help your data analysis and also make easy the use of many libraries in R.

```{r}
library(readr)# read cvs
library(readxl) #read xls
library(dplyr)
library(knitr)      # web widget
library(tidyverse)  # data manipulation
library(data.table) # fast file reading
library(kableExtra) # nice table html formatting 
library(gridExtra)  # arranging ggplot in grid
library(caTools)    # split 
library(plotrix)
library(MASS)
```

### Import data

Data used in this material are downloaded from this link: https://www.kaggle.com/sonujha090/bank-marketing

```{r}
data_bank
```

```{r}
data_bank_tibble<-tibble(data_bank)
head(data_bank_tibble)

```

Looking at the dataset
```{r}
# View(data_bank)# view dataset
head(data_bank)# head of rows
summary(data_bank)# summary statistics
str(data_bank) # structure of dataset
nrow(data_bank)# number of rows
ncol(data_bank)# number of columns
dim(data_bank)# dimensions of dataframe
```

### If we want to code binary data 

Lets try it in "loan" variable with categories (yes, no). We are going to use the ifelse() function.
```{r}
table(data_bank$housing)
table(data_bank$marital,data_bank$job)# contigency table

# Now let see the difference
data_bank$housing= ifelse(data_bank$housing=='yes',1,0)

head(data_bank$housing,10)
table(data_bank$housing)
```
## Train and test data
Let's work with our first format dataset. We will import it again here.

```{r}

n.train<-0.8*nrow(data_bank)# number of observ in training set
train<-head(data_bank,n.train)# you may use sample
nrow(train) # length of training data

n.test<-0.2*nrow(data_bank)
test<-tail(data_bank,n.test)
nrow(test) # length of testing data
```


### Check for "missing values, NA"
```{r}
sum(!complete.cases(data_bank)) # it will turn 0 because this dataset is cleaned from NA-s
```
### Check for dublicates in rows 
```{r}
sum(duplicated(data_bank))
# it will return the number of duplicated , here it will return 0 because we have cleaned the duplicate
```

### Descriptive statistics

```{r}
summary(data_bank_tibble)
summary(data_bank_tibble$age)# only for age variable
summary(data_bank_tibble[,3:6])# only for variables in columns 3 up to 6
```
### Data Wrangling 
#### dplyr package

```{r}
library(dplyr)
```

#### filter()
filter() is a function in dplyr that takes logical expressions and returns the rows for which all are TRUE.


```{r}
# filter individuals of age less than 25 years old
filter(data_bank, age < 25) 

# filter individuals housing == yes
filter(data_bank, housing== "yes")

# individuals with profession management and age =30 years old
filter(data_bank, job == "management", age == 30)

# individuals with profession management and age less than 25 years old 
filter(data_bank, job == "management", age < 25)

# we want to display in job the retired and management professions
filter(data_bank, job %in% c("retired", "management")) 
```

### Exercise
a. What was the average age of "management" professionals housing "yes".

Hint: you can do this in 2 steps by assigning a variable and then using the mean() function.

###### Solution. a
```{r}
av.age <- filter(data_bank, job == "management", housing== "yes")  
mean(av.age$age)  
```


b. What was the average balance of the loan for secondary education individuals?  


### select() 

We use select() to subset the data on variables or columns.

We can select multiple columns with a comma, after we specify the data frame (data_bank).

#### The logic of select():

   select(df, A, B ,C): Select the variables A, B and C from df dataset.
   select(df, A:C) : Select all variables from A to C from df dataset.
   select(df, -C, -A): Exclude C from the dataset from df dataset.	

Sometimes you may notice an ERROR after applying function select(); this is likely that you are either using a package besides dplyr that also has a select() function or you just forgot to load the dplyr package with library(dplyr). If this happens you may follow the syntax below (data_bank %>% dplyr::select(age))

```{r}
library(dplyr)
# data_bank %>% select(education) # Error in select(., education) : unused argument (education)
```
We can use this syntax of obtaining the results with select()
```{r}
head(select(data_bank,age))
head(data_bank %>% dplyr::select(age))

data_bank %>% dplyr::select(education)

# or select 1st variable (age)
head(data_bank[,c(1,5,10)]) 

data_bank %>% dplyr::select(age, marital, duration) 
```
 
#### De-select variables  
We can also use - (minus) to deselect columns. 
The code below will de-select from our dataset the variables age and marital.

```{r}
D2<-data_bank %>% dplyr::select(-age, -marital) 
head(D2)
```

#### We can use the pipe to chain those two operations together.

We want to filter for profession (job - variable) the "management" and from these individuals to show their "education" and "loan" variable.

#### Exercise
How can you do it for individuals job "management" status to show their "age" and "balance"

```{r}
data_bank.1 <- data_bank %>% 
  filter(job == "management") %>% 
  dplyr::select(education,loan) 

data_bank.1
```

### mutate() adds new variables to dataframe

Let suppose we want to add a variable combining two or more existing variables. Or we want to add a new variable from an existing vector which is not part of the dataframe.

Let's suppose I want to see the day balance. (balance/duration).

Adding a new variable from two existing variables of the dataframe, named "day.balance"
```{r}
D3<-data_bank %>%
  mutate(day.balance = round(balance/duration,2))
D3
```

I can add from another existing variable outside the dataset.
```{r}
X<-sample(c("a","b","c"),length(data_bank$age),replace=TRUE)
X1<-sample(c("a","b","c"),nrow(data_bank),replace=TRUE)

length(X)
data_bank %>% mutate(X)
```

### group_by() operates on groups

summarize() will actually only keep the columns that are grouped_by or summarized. 

ungroup() removes the grouping and it’s good to get in the habit of using it after a group_by().
```{r}
data_bank %>%
  filter(age == 30) %>%
  group_by(marital) %>%
  summarize(mean(duration)) %>%
  ungroup()
```

### arrange()

arrange() function is ordered alphabetically from A to Z.

```{r}
# arrange by job
data_bank %>%
   group_by(marital,job) %>%
  summarize(mean(duration)) %>%
  arrange(job)

# arrange by marital status
 data_bank %>%
   group_by(marital,job) %>%
  summarize(mean(duration)) %>%
  arrange(marital) 
```
Try to combine by yourself the above functions. 

### Exercise
What is the maximum duration for all jobs at age 30?

### Exercise
Try to understand what the below command will give you as an output?
```{r}
library(tidyverse) ## install.packages('tidyverse')
## summarize
data_bank.2 <- data_bank %>% 
  dplyr::select(-contact, -poutcome) %>% 
  dplyr::group_by(marital) %>%
  dplyr::mutate(day.balance = round(balance/duration,2)) %>%
  dplyr::summarize(min_day.balance = min(day.balance)) %>%
  dplyr::ungroup() 

```




### February 2022
#### Eralda Gjika (Dhamo)

